---
title: "Jens TRY Update - Small Trials"
author: "Gijsbert Werner, University of Oxford"
date: "4/28/2019"
output: html_document
---
```{r,include=F}
#Load libraries
library(ape)
library(phytools)
library(dplyr)
library(data.table)
library(RColorBrewer)
library(viridis)
library(diversitree)
library(corHMM)

```

This document contains a first pass at generating nice overview figures for the TRY Update paper (email Jens April 2019). This is a very general first attempt to highlight some potential options, so disregard all the specifics of colouring scheme, font, size etc. etc. These can all be adapted later, for instance to match colouring schemes used elsewhere in the paper. 

# Quick data check

```{r, include = F}
#Read in data directly from the zipped file
dat<-as.data.table(
  read.csv(unz("./Data/ALLMB_Consolidated_FleshedGenera_TRY_Traits.csv.zip",
             "ALLMB_Consolidated_FleshedGenera_TRY_Traits.csv")))
tree<-read.tree("./Data/v0.1/ALLMB.tre")
```
Using Jens second data file (email 26/04/2019), let's have a look a the data. 
```{r}
str(dat)
```
Ok, this looks sensible. Let's have a look at the phylogeny. 

```{r}
tree
```

Yes, this makes sense too. The number of tips in the phylogeny is exactly the number of rows in the data frame. It's seems that the column that matches them to each other is ALLMB_OrigSpecies. Check this: 

```{r}
dat$match_col<-gsub(pattern=" ","_",dat$ALLMB_OrigSpecies)
length(which(dat$match_col %in% tree$tip.label))
length(which(tree$tip.label %in% dat$match_col))
```

This looks mostly good, but there are 142 species without two-way match. I think it's spelling things. I will drop them. There's one more thing that is a bit odd. If we look at the TRY_AccSpeciesID column which indicates the TRY species this has been matched to we see quite a lot of duplicates, indicating that multiple phylogeny tips are matched to the smae TRY species. This makes some sense given that quite some of the mega-phylogeny tips are spelling error duplicates, but let's have a look at the numbers. 

```{r}
length(which(duplicated(dat$TRY_AccSpeciesID)))
length(unique(dat$TRY_AccSpeciesID))
```

So, this means that over 140k rows / tips are non-unique in the data frame (and in the phylogeny?), and almost 216k are unique. I am not sure if this is worrying. I suppose it means that lots of things in the phylogeny that are represented as distinct tips should really be only a single one. If that interpretation is accurate, we are really looking at an ~210k species phylogeny instead of a 350k one (still pretty impressive!). 

Given that our purpose is visualisation, and not 'real' analysis this will probably be fine to ignore. Another option would be prune the phylogeny to only include the unique 216k species. This would be pretty easy to do based on what we have. Let me know what you think, Jens. For now, I will just ignore this. 

# Initial Trial Visualisation

Essentially, we have (i) a bunch of different binary data, e.g. Any data, specific traits (SLA), the six Diaz traits and (ii) quantitative, i.e. number of data. The latter can also represent the former, given that the absence of data just means a 0 on a continous scale. 

I will randomly selected 75 tips/rows to higlight some combinations of options. The reason I am not immediately going into the full 350k is that rendering a figure of that size can actually takes a while (sometimes hours, even on my relatively fast computer), so I want to have a bit of an idea first of where we are heading. 

A few potential directions follow. 

```{r, include=F}

#Let's make some small dataset for trial code.
set.seed(01865)
small_dat<-dat[sample(nrow(dat),size=75),]
small_tree<-drop.tip(tree,
                     tree$tip.label[!tree$tip.label %in% small_dat$match_col])
```

## Binary Data

An option is to highlight presence/absence of one or multiple binary (or categorical) traits using colour-coded concentric rings around the phylogeny. This is what we did in Figure 2 in our PNAS paper last year (have a look). I will now, for the 75 random species phylogeny, do this to visualise (1) Any data present, (2) Six Diaz traits and (3) SLA (just a random trait), in a single figure (Figure 1; big 300k version attached as separate pdf).

```{r,echo=F}
binary_traits<-small_dat %>% select(Number.of.Traits,SLA,All.six.traits..Diaz.et.al.2016.)
names(binary_traits)[1]<-"Presence"
names(binary_traits)[3]<-"All.six"
binary_traits$Presence<-ifelse(is.na(binary_traits$Presence),0,1)
binary_traits$All.six<-ifelse(is.na(binary_traits$All.six),0,1)
binary_traits$SLA<-ifelse(is.na(binary_traits$SLA),0,1)
rownames(binary_traits)<-small_dat$match_col
trait.plot(tree = small_tree,dat = binary_traits,cols = list(Presence=c("gray90","red"),
                                                             SLA=c("gray90","green"),
                                                             All.six=c("gray90","blue")),
           legend=T,cex.lab=0.0001)
```

Ok, nice. From the inside to the outside we see data presence in Red, then SLA presence in green and then Diaz all six traits prence (just a single species). 

In principle, you can keep expanding the number of rings, we can for instance look at all the six traits of interest individually (Figure 2):

```{r,echo=F}
binary_traits2<-small_dat %>% select(Leaf.Area,SLA,Leaf.Nitrogen.Content.Per.Dry.Mass,Seed.Dry.Mass,Plant.Height,Stem.Specific.Density..SSD.)
binary_traits2[is.na(binary_traits2)]<-0
rownames(binary_traits2)<-small_dat$match_col

trait.plot(tree = small_tree,dat = binary_traits2,cols = list(Leaf.Area=c("gray90","red"),
                                                             SLA=c("gray90","pink"),
                                                             Leaf.Nitrogen.Content.Per.Dry.Mass=c("gray90","green"),
                                                             Seed.Dry.Mass=c("gray90","blue"),
                                                             Plant.Height=c("gray90","purple"),
                                                             Stem.Specific.Density..SSD.=c("gray90","orange")),
           legend=T,cex.lab=0.0001,font.lab = 0.35)
```

At some point, this stops working though, because things will get messy with two many rings. Six is probably pushing it, but it might work. We would have to look at what it looks like with the full phylogeny to see if it works. 

## Continuous Data

Let's have a quick look at the distribution of the number of traits in this species set:

```{r}
summary(small_dat$Number.of.Traits)
```

There is lots of variation among those species where we have data in the number of traits we have data for. The median is 2, the maximum is 274 (!), i.e. two orders of magnitude higher. This can make visualisations a bit difficult, because it makes it really difficult to both make variation in the lower numbers (1-10), which we are probably most interested in, and big outliers visual. For instance, let's have a look (Figure 3).

```{r,echo=F}
#Prep vector
small_trait_num<-small_dat$Number.of.Traits
names(small_trait_num)<-small_dat$match_col
small_trait_num[is.na(small_trait_num)]<-0 #Because NA means 0 traits here
#Visualise this in two ways
plotTree.wBars(small_tree,x = small_trait_num,border="white",type="fan")
```

Perhaps unsurprisingly, the only species where we have all the six Diaz traits, is the one with the 274 traits available. It's Festuca ovina, probably a well-studied/common species. 

An option to make the variation in trait number slightly more visually perceptible is by taking the natural log (Figure 4):

```{r,echo=F}
#Take natural log of values
small_trait_num_log<-ifelse(small_trait_num>0,log(small_trait_num),0)
plotTree.wBars(small_tree,x = small_trait_num_log,border="white",type="fan")
```

This of course compresses the really big species numbers, and also make immediately obvious what species don't have any data (i.e. the ones without bars). In that sense, it basically tells us the same as the inner ring of Figure 1, but also a bit more. We can improve it even further (Figure 5, Second big 300k separate pdf figure):

```{r,echo=F}
plotTree.wBars(small_tree,x = small_trait_num_log,border="white",type="fan",
               col = viridis(100)[as.numeric(cut(small_trait_num_log[match(small_tree$tip.label,names(small_trait_num_log))],
                                              breaks = 100))])
add.color.bar(100,viridis(100),title = "Log of trait #",prompt = F,
              lims = c(min(small_trait_num_log),max(small_trait_num_log)),
              x=-185,y=-185)
```

Now, I have coloured the bars with a colour indicating the gradient of trait numers (ranging up to 5.6, because we are loooking at logs and e^5.6 = 274). So, here, blueish is low data availability, yellower higher. 

One more addition (Figure 6):

```{r,echo=F}
small_tree_rec_num_log<-contMap(tree = small_tree,x = small_trait_num_log,outline = F,
                            type="fan",col="white",fsize=c(0,1),plot=F)
small_tree_rec_num_log_viridis<-setMap(small_tree_rec_num_log,viridis(100))
#And with a coloured branch and bars
plotTree.wBars(tree=small_tree_rec_num_log_viridis$tree,x = small_trait_num_log,
               method = "plotSimmap",
               colors=small_tree_rec_num_log_viridis$cols,
               fsize=c(0,1),type="fan",border="white",
               col = viridis(100)[as.numeric(cut(small_trait_num_log[match(small_tree$tip.label,names(small_trait_num_log))],
                                                 breaks = 100))])
```

Now, we have a quantitative ancestral state reconstruction on the phylogeny branches. 

No suprises, I guess, but what do the branch colours mean? Some notion of data presentness, I suppose, which is perhaps not really a plant trait. If we want to go down this route, we have to think a bit more about this (also about how to treat the absences, which are now just the lowest point on continuous range - this may or may not be ok. An alternative option is to drop the absence, and only do this type of anlaysis on the species where we have data: the figure will then show something else, i.e. for those species where we have data, how much?)

## Ordinal Data

A third strategy is to treat the number of traits as an ordinal variable, i.e. distribute it in bings. How about this (Figure 7):

```{r}
####Try this as category ranges:
#E.g. colours as (1-5,5-10,10-100,>100) via diversitree plot
dat_num_categories<-small_dat %>% select(Presence=Number.of.Traits) 
dat_num_categories$Presence<-ifelse(is.na(dat_num_categories$Presence),0,dat_num_categories$Presence)
dat_num_categories <- dat_num_categories  %>% 
  mutate(Presence=cut(Presence,
                      breaks = c(-Inf,1,6,11,101,Inf),
                      labels = c("Absence","# 1-5","# 6-10","# 11-100","# > 100"),right=F))
rownames(dat_num_categories)<-small_dat$match_col

trait.plot(tree = small_tree,dat = dat_num_categories,cols = list(Presence=c("gray90","#fecc5c","#fd8d3c","#f03b20","#bd0026")),
           legend=T,cex.lab=0.0001,font.lab = 0.5)
```

Ok, so now we have five bins, data absence (gray) and then in colours 1-5 traits, 6-10 traits, 11-100 and more than hundred. So basically, poorly studies species, reasonably well studied species, well studies species, and extraordinarily well studies species. Of course, many other splits and bins are possible. 

# Thoughts and conclusions

There is lots and lots of things we can do, some thoughts/considerations:

- Many of the things above would could be combined with each / other things. For instance, the figures 1+2 could be combined with an ancestral state reconstruction of one of the underlying binary variables, plotting on the phylogeny with coloured branches (e.g. our PNAS paper). Many other combinations of the main elements from above (bars, coloured branches, coloured rings around the phylogeny). 
- None of the above figures have any indication of taxonomy, which people will want to have a rough idea about. Some options to this are 1. coloured branches indicating main groups (e.g. angiosperms, gymnosperms), 2. little manual drawings at the outside of the phylogeny indicating main figures. 3. Textual labels outside the phylogeny 4. A combination of labels and figures (see our PNAS paper again), 5. Concentric rings with colours (see our Nature Communications paper from 5 years ago).
- Some of the things above may not scale well to 300k species. We should probably keep it simple at that scale. 

Ultimately, the choice of figures is primarily a matter of what we want to show, and what the funciton of the figure(s) in the paper will be. 
I think that at this stage, the most productive thing is therefore to think a bit deeper about what we want the figure /figures to show, and ask how we can best approach this from there. What is the specific message of the figure in the manuscript? How much space do we have? One big figure? A few big figures? One big and a few small? 

I think it's best, Jens, if you think about questions like this a bit more. We might even have a Skype-chat. If you send me a draft of the ms (do you have it already?), I could also think along those lines. For now, my two cents are:

- A version of the first figure, with as main message the simple patterns of presence / absence, potentially with coloured branches indicating either number of traits, specific traits of interest or some form of taxonomy is likely the key most interesting contribution. Basically, the big version of Figure 1, with whistles and bells, and made nice.  
- In addition, something with number of traits (for all species, or only for those species with at least one trait present): here I like the coloured barplots around the phylogeny, potentially with an ancestral state reconstruction, 
- Potential second addition: something with the six variables of interest, either individually, or as a single (binary variable). Basically, the big version of figure 
This way, you go from broadest (i.e. any presence?) to more specific (how much data) to very specific (which data for specific traits).

Lastly, of course, subclades of interest are always an option. I have no good insight here what is useful yet. Let's save for later. 



